---
title: "Potential In London"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PotentialInLondon}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HodgePotentialHumanFlow)
```

# Summary
This vignette shows the scalar potential of a commuting flow in London as a example.

# Dataset: origin-destination matrix in London
We use a trip dataset from home to workplace in 2011 obtained from [UK Data Service](http://www.nomisweb.co.uk/census/2011/wu03EW).
The OD matrix denotes the number of commuters aggregated by the middle layer super output area (MSOA) in the census 2011.
The dataset covers the MSOAs in England and Wales.
Here, we selected the trips only among MSOAs in Greater London.
```{r}
od = read.csv("london.csv") 
head(od)
```

# Hodge-Kodaira decomposition
Here, we apply Hodge-Kodaira decomposition, in which human flow is uniquely decomposed into a potential-driven (gradient) flow and a curl flow (see our paper for the details).

First, we consider the net flow of movement from a given  OD matrix $M$:
$$
  A = M - M^{\intercal}, \label{eq:netflow}
$$
where $M^{\intercal}$ denotes the transpose of $M$.
Matrix $A$ is skew-symmetric, that is, $A_{ij} = - A_{ji}$,
and is possibly described by combinatorial gradient of a potential $s$, given by
$$
  (\text{grad}\, s)(i, j) =  s_j-s_i.
$$

Then, we define an optimisation problem for potential $s$:
\begin{equation}
  \min_s  \lVert   \text{grad}\ s - A \rVert_2 = \min_s \left[ \sum_{i,j} \left[ (s_j - s_i) - A_{ij} \right]^2 \right]. \label{eq:optimization_problem}
\end{equation}
According to the combinatorial Hodge theory, the space of net flow $\mathcal{A}$ is orthogonally decomposed into two subspaces:
\begin{equation} 
  \mathcal{A}  = \text{im}(\text{grad})  \oplus \text{im}(\text{curl}^*),
\end{equation}
where $\text{curl}$ is the combinatorial curl operator and $\text{curl}^*$ is its adjoint operator.
Thus, the optimisation problem is equivalent to an $l_2$-projection of $A$ onto im(grad),
and the minimal norm solution is simply given by
\begin{equation}
  s_i = -\frac{1}{N} \text{div} A = - \frac{1}{N} \sum_{j=1}^N A_{ij}, \label{eq:potential}
\end{equation}
where $s_i$ is the potential at the $i$th location and $N$ is the number of locations. 
It is noted that $s_i$ is negative potential ($s_i=-V_i$).
This means that we observe more trips from a place with lower potential to another place with higher potential.

Using this package, the potential is obtained by the following function.
```{r}
s = HodgePotentialHumanFlow::scalar_potential(od)
head(s)
```

# Plot the obtained potential landscape 
The shapefile of the MSOAs is obtained from [Office for National Statistics](https://data.gov.uk/dataset/2cf1f346-2f74-4c06-bd4b-30d7e4df5ae7/middle-layer-super-output-area-msoa-boundaries)).

```{r}
library(sf)
shape = st_read("london.geojson") 
merged = merge(shape, s, all.x=T, by.x="msoa11cd", by.y="zone")
head(merged)
```

Plot the obtained potential landscape using ```sf``` package.
```{r fig.asp = 0.8, fig.width = 7, out.width = "100%"}
library(ggplot2)
library(viridis)
library(scales)
ggplot(data=merged, aes(fill=potential)) + geom_sf(color="black", lwd=0.2)  + scale_fill_viridis_c(option="viridis", trans=modulus_trans(0)) + theme_bw() + theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank(),panel.grid.major = element_line(colour = "transparent"))
```
